Abstract Spatial Specification of Stencil Computations

### 2016-07-03 paper.tex
related work on PLDI'16 paper, and examples

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}

@@

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

@@

\newcommand{\stenRefl}[1]{\term{reflexive} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenReflS}[1]{\term{refl} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}

@@

\include{results}
\include{corpussize}

\title{Verification of Stencil Computations through Spatial Specifications}
\authorinfo{Dominic Orchard \and Mistral Contrastin
\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{modMult},] \; \nonterm{refRegion} \\
\nonterm{refRegion} ::= & [\nonterm{modRefl},] \; \nonterm{region}\\
\nonterm{modMult} ::= & \term{readOnce} \\
\nonterm{modRefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::=
        & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenRefl{\dimTy}  \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Note that there are a number of intermediate assignments over which
the whole specification ranges.

@@

To save space we use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, and \term{cen} for \term{reflexive},
\term{forward}, \term{backward}, and \term{centered} regions respectively.
The equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

@@

Figure~\ref{fig:region-model} outlines the core rules for regions
via the interpretation function $\interp{-}$.  We explain the
intermediate notations and definitions:

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $\infty$, \eg{},
$\singleEntry{1}{2} = \vtwoh{1}{\infty}$.
\end{definition}

\begin{definition}A \emph{zero-entry vector} of size $n$, denoted
$\zeroEntry{r}{n}$, is a vector where the $r^{th}$ entry is $0$ and all others
are $\infty$, \eg{}, $\zeroEntry{1}{2} = \vtwoh{0}{\infty}$.
\end{definition}

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{alignat*}{2}
& dims(M_m,N_n) =
  \bigg\{ x \; \bigg| \;
    \parbox{4.1cm}{
      $x \in \{1, \ldots, m\} \wedge M_x \neq \infty \; \vee \;$ \\
      $x \in \{1, \ldots, n\} \wedge N_x \neq \infty$
                  } \bigg\} \\
& M_n \otimes N_n =
  \Bigg\{x \; \Bigg| \;
    \parbox{4.8cm}{$i \in \{1, \ldots, 2^n\},$
                   $(u, v) \in M_n \times N_n,$ \\
                   $x = (u \bowtie v)_i,$ \\
                   $\forall j \in dims(M_n,N_n) (x_j \neq \infty) $
                  } \Bigg\}
\end{alignat*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
\{\infty\}$. In pairwise permutation $\infty$

@@

\begin{figure}
\input{region-model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Set model of regions}
\label{fig:region-model}
\end{figure}

\begin{figure}
\input{spatial-model.tex}
\caption{Model of specifications}
\label{fig:spatial-model}
\end{figure}

\begin{figure}
\input{consistency.tex}
\caption{Consistency function for a model and a set of indices}
\label{fig:spatial-model}
\end{figure}

@@

Informally, the procedure proceeds as follows. First, relative
indices are turned into unit regions in $n$-dimensions. For
example, for the five point stencil there are five $1 \times 1$ regions,
called \emph{spans}, which are essentially $0$-dimensional:
%
\begin{center}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\end{center}
%
Each dimension is traversed, coalescing consecutive spans
which vary in $m$-dimensions into coalesced spans which vary in
$m+1$-dimensions. In this example, the $0$-dimensional spans
 become $1$-dimensional spans (rows/columns):
%
\begin{align*}
\begin{array}{cc}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.40) rectangle (1.5,0.90);
\draw[thick] (1,1.60) rectangle (1.5,2.10);
\end{tikzpicture}
\qquad
&
\qquad
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1.6,1) rectangle (2.1,1.5);
\draw[thick] (0.4,1) rectangle (0.9,1.5);
\draw[thick] (1,0.5) rectangle (1.5,2);
\end{tikzpicture}
\\
\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
\end{array}
\end{align*}
%
Any spans that are contained with any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another).
\begin{center}
\begin{tikzpicture}
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,0.5) rectangle (1.5,2);
\node at (2,1.22) {+};
\draw[thick] (2.5,1) rectangle (4,1.5);
%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
\end{tikzpicture}
\end{center}
%
The procedure is then iterated till a fixed point is reached. In this
example this is reached in the first step. These final spans
go on to form the basis of the spatial stencil specification, where
each span is a region product \texttt{*} and multiple spans are
combined with the region sum \texttt{+}.

@@

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be unpermuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

\item Sort each permutation set into an ordered list, based on the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalescees contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\textsf{regions}(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.

@@

\section{Evaluation}
\label{sec:evaluation}
We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.

\subsection{Software Corpus}
\begin{figure}
\begin{tabular}{|l|l|l|l|}
\hline
Package       & Files & Lines of code & Files analyzed \\
\hline
UM            & \umFiles & \umLoC & \umparseOk  \\
E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC & \ethreemgeaparseOk  \\
BLAS          & \blasFiles   & \blasLoC & \blasparseOk        \\
Hybrid4       & \hybridfourFiles    & \hybridfourLoC & \hybridfourparseOk         \\
GEOS-Chem     & \geoschemFiles   & \geoschemLoC & \geoschemparseOk       \\
Navier        & \navierFiles     & \navierLoC & \navierparseOk           \\
CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC & \computationalphysicstwoparseOk           \\
\hline
\textbf{Total}& \overallFiles  & \overallLoC   & \overallparseOk  \\
\hline 
\end{tabular}
\caption{Summary of software packages used for evaluation\label{fig:corpus}}
\end{figure}

>>>>

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}

\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}

@@

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

@@

\newcommand{\stenFwd}[2]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwd}[2]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCen}[2]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}

@@

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}

@@

\title{Abstract Spatial Specification of Stencil Computations}
\authorinfo{Dominic Orchard \and Mistral Contrastin
\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{D}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{specTop} \;
                        \texttt{::} \; v
  \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[1em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{specTop} ::= & [\nonterm{approxMod},] \; \nonterm{spec} \\
\nonterm{spec} ::= & [\nonterm{mod}] \; [\nonterm{region}] \\
\nonterm{mod} ::= & [\nonterm{modR},] \; [\nonterm{modRefl},] \; [\nonterm{modIrefl},] \\
\nonterm{modR} ::= &
 \term{readOnce} \\
\nonterm{modRefl} ::= & \term{reflexive} \, (\term{dims=}\dimTy \{\term{,} \; \dimTy\}) \\
\nonterm{modIrrefl} ::= & \term{irreflexive} \, (\term{dims=}\dimTy \{\term{,}
          \dimTy\}) \\
\nonterm{approxMod} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \stenFwd{\mathbb{N}_{>0}}{\dimTy}  \\
\mid \; & \stenBwd{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \stenCen{\mathbb{N}_{>0}}{\dimTy} \\
\mid \; & \nonterm{region} \; \term{+} \; \nonterm{region} \\
\mid \; & \nonterm{region} \; \term{*} \; \nonterm{region} \\
\mid \; & \nonterm{rvar}  \\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\dimTy ::= \; & \mathbb{N}_{>0} \\
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

@@

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Such a stencil computation is a common image convolution (for example
for edge detection). 
Note that there are a number of intermediate assignments over which
the whole specification ranges.

As another example, the following code and specification
implements the \emph{Roberts cross}
edge-detection convolution~\cite{davis1975survey}:
%%
\begin{ExmVerbatim}
do j=0, jmax-1
    do i=0, imax-1
\sten{      != region r1 = forward(depth=1, dim=1)}
\sten{      != region r2 = forward(depth=1, dim=2)}
\sten{      != stencil readOnce, r1*r2 :: a}
      x = a(i,j) - a(i+1,j+1)
      y = a(i+1,j) - a(i,j+1)
    b(i,j) = sqrt(x*x+y*y)
   end do
end do
\end{ExmVerbatim}

@@

To save space we use abbreviations
\term{fwd}, \term{bwd}, and \term{cen} for \term{forward},
\term{backward}, and \term{centered} regions respectively.  The
equations reveal something of their semantics.  The first six
equations (with labels of the form $\ldots \texttt{+} \ldots$)
explain the behaviour of overlapping regions with sum
\term{+}. The next six equations explain the algebraic behaviour of
\term{+} and \term{*}.  Together, we see that \term{+} is an
associative, commutative, idempotent binary operator that distributes
with \term{*}, which is associative and commutative. This distribution
of \term{*} over \term{+} is key as it is used internally to give a
normal form for stencil specifications akin to Disjunctive Normal Form
(DNF) (where \term{+} is taken as ``disjunction'' and \term{*} as the
``conjunction''). We revisit this normalisation in Section~\ref{}.

@@

Figure~\ref{fig:model} outlines the core rules for regions
via the interpretation function $\interp{-}$.  We explain the
intermediate notations and definitions:

\begin{definition}A \emph{single-entry vector} of size $n$, denoted
$\singleEntry{r}{n}$, is a vector
where the $r^{th}$ entry is $1$ and all others are $0$, \eg{},
$\singleEntry{1}{2} = \vtwoh{1}{0}$.
\end{definition}

\begin{definition}The product of two models $\otimes$ is defined:
%%
\begin{align*}
M \otimes N & =
  \{(u \bowtie v)_i \mid i \in \{1, \ldots, 2^n\}, (u, v) \in M \times N\}
\end{align*}
The $\bowtie$ operation, which we call \emph{pairwise permutation}
(described below), constructs a $2^n \times n$ matrix ($2^n$-vector of
$n$-vectors). Using this, the $\otimes$ operation constructs the set of every
pairwise permutation for every pair $(u, v)$ of $n$-vectors in the Cartesian
product of $M \times N$.

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain
$\mathbb{Z} \cup \{\infty\}$.
For addition and multiplication, we take
 $\infty$ to be an absorbing element. Thus, if either
vector in a pairwise permutation contains $\infty$ in
the $j^{th}$ entry, then $j^{th}$ entry of every pairwise
permutation is $\infty$, \eg{},
%%
\begin{equation*}
\vtwoh{0}{1} \bowtie \vtwoh{2}{\infty} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{cc}
0 & \infty \\
0 & \infty \\
2 & \infty \\
2 & \infty
\end{array}
\right]
\end{equation*}

@@

\begin{figure}
\input{model.tex}
%\textit{$a^m \in A$ means there are $n$ copies of $a$ in
%  the multi-set $A$}
\caption{Set model of specifications}
\label{fig:model}
\end{figure}

@@

Informally, the procedure proceeds as follows. First, relative
indices are turned into unit regions in $n$-dimensions. For
example, for the five point stencil there are five $1 \times 1$ regions,
called \emph{spans}, which are essentially $0$-dimensional:
%
\begin{center}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\end{center}
%
Each dimension is traversed, coalescing consecutive spans
which vary in $m$-dimensions into coalesced spans which vary in
$m+1$-dimensions. In this example, the $0$-dimensional spans
 become $1$-dimensional spans (rows/columns):
%
\begin{align*}
\begin{array}{cc}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\qquad
&
\qquad
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,0.5) rectangle (1.5,2);
\end{tikzpicture}
\\
\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
\end{array}
\end{align*}
%
Any spans that are contained with any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another).
\begin{center}
\begin{tikzpicture}
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,0.5) rectangle (1.5,2);
\node at (2,1.22) {+};
\draw[thick] (2.5,1) rectangle (4,1.5);
%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
\end{tikzpicture}
\end{center}
%
The procedure is then iterated till a fixed point is reached. In this
example this is reached in the first step. These final spans
go on to form the basis of the spatial stencil specification, where
each span is a region product \texttt{*} and multiple spans are
combined with the region sum \texttt{+}.

@@

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be unpermuted later.
%
\[
P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{-1}{0}{-1}{0}
\stwo{0}{1}{0}{1}
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0}
 \stwo{0}{1}{0}{1}
 \stwo{0}{-1}{0}{-1}
 \stwo{1}{0}{1}{0}
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

\item Sort each permutation set into an ordered list, based on the ordering:
\[
\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i
\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
\]
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalescees contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
 \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}
\stwo{0}{1}{0}{1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{-1}{0}{1}
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\textsf{regions}(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.

@@

\section{Evaluation}
\label{sec:evaluation}
